Constraints on Cosmic-Ray Particle Populations, Magnetic Fields, and Dark Matter from VHE Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS

### 2011-09-20 ms.tex
rewrote abstract

\begin{abstract}
Theoretical models of galaxy clusters suggest they may be sources of very high-energy (VHE)
gamma-ray emission. Observations of radio halos and relics in clusters are indicative efficient
acceleration processes and the gamma-ray emission is the result of highly relativistic
cosmic-ray protons and ions interacting with intracluster gas and magnetic fields. Reported
here are VHE gamma-ray observations of the Coma cluster of galaxies with the VERITAS array
of imaging Cherenkov telescopes, with complementing Fermi Gamma-ray Space Telescope
observations at GeV energies. No significant gamma-ray emission from the Coma cluster
was detected. Flux upper limits at the 99\% confidence level were measured to be on the order of $2-5\times 10^{-8}$ ph. m$^{-2}$ s$^{-1}$ (VERITAS, $>$220 GeV) and
$\sim 3\times 10^{-6}$ ph. m$^{-2}$ s$^{-1}$ ({\em Fermi}, 1-3 GeV), respectively. The limits are
two orders of magnitude higher than model-predicted minimum gamma-ray fluxes, which
assumes the radio-emitting electrons to be all secondaries from cosmic-ray interactions.

The gamma-ray flux upper limits have been used to constrain the populations of cosmic-ray
particles and magnetic fields and in the Coma cluster of galaxies. Using a simplified analytical
approach, the relative cosmic-ray pressure, $X_{\mathrm{CR}}=\expval{P_{\CR}}/\expval{P_{\rmn{th}}}$,
is constrained to be smaller than 20\% from VERITAS data and smaller than 4\% from
{\em Fermi} data. 
The observations imply a lower limit on
the intra-cluster magnetic field on the order of a $\mu$G, depending on the magnetic decline
and gamma-ray spectral index. Finally, a constraint on the thermally averaged product of the total self-annihilation
cross section and the relative velocity of the dark-matter particles, $\expval{\sigma v}$,
is derived and found to be $\lesssim 10^{-21}$ cm$^{3}$ s$^{-1}$ for $m_{\chi}\gtrsim 1000$
GeV/c$^{2}$.
\end{abstract}

@@

The ICM is a hot, $T\sim 10^{8}$ K, plasma emitting thermal bremsstrahlung in the soft
X-rays regime \citep[see, e.g.,][]{article:Petrosian:2001}. This plasma has been heated
primarily through collisionless structure-formation shocks that form as a result of the
hierarchical merging and accretion processes. Such shocks and turbulence in the ICM gas and
intra-cluster magnetic fields also provide a means to efficiently accelerate particles \citep[see,
e.g.,][]{article:ColafrancescoBlasi:1998, article:Ryu_etal:2003}. Many clusters do indeed feature
mega-parsec scale halos of non-thermal radio emission, indicative of a population of relativistic
electrons and positrons (leptons) and magnetic fields permeating the ICM
\citep{article:Cassano_etal:2010}. The two theories competing to explain the radio halo
are the ``hadronic model'', in which the radio-emitting electrons are produced in inelastic
collisions of cosmic-ray (CR)  ions with thermal gas of the ICM \citep{article:Dennison:1980},
%\citep[][and references therein]{article:EnsslinPfrommerMiniatiSubramanian:2011}
and the ``re-acceleration model'', in which during states of strong ICM turbulence, e.g. after a
cluster merger, interactions of a long-lived pool of 100-MeV electrons with plasma waves might
be efficient enough to accelerate them to energies ($\sim 10$ GeV) sufficient to produce observable
radio emisson \citep[][and references therein]{article:BrunettiLazarian:2010}. Recent observations of
possibly non-thermal emission from clusters in the the extreme ultraviolet
\citep[EUV; ][]{article:SarazinLieu:1998} and hard X-rays \citep{article:RephaeliGruber:2002, article:Fusco-Femiano_etal:2004, article:Eckert_etal:2007} may provide further indications of
relativistic particle populations in clusters, although the interpretation of these observations as non-thermal diffuse emission have been disputed on the basis of more sensitive observations
\citep[see, e.g.,][]{article:Ajello_etal:2009, article:Ajello_etal:2010, article:Wik_etal:2009}.

@@

\section{Gamma Ray Emission from Cosmic Rays}
We decided to adopt a multi-faceted approach to constrain the CR-to-thermal pressure distribution
in the Coma cluster using the upper limits dereived from VERITAS and \emph{Fermi}-LAT data in this
work. This approach includes (1) a simplified multi-frequency analytical approach that assumes a
constant CR-to-thermal energy density and a power-law spectrum in momentum, (2) an analytic model
derived from cosmological hydrodynamical simulations of the formation of galaxy clusters, and (3)
an approach that uses the
observed luminosity of the giant radio halo in Coma to place a lower limit on the expected
$\gamma$-ray flux in the hadronic model -- where the radio-emitting electrons are secondaries from
CR interactions and which is independent of the magnetic field distribution.  This translates into a
minimum CR pressure which, if challenged by tight gamma-ray limits/detections, enables scrutinizing
the hadronic interaction model of the formation of giant radio halos. Alternatively, by realizing a
spatial CR distribution that is just consitent with the flux upper limits and requiring the model to
match the observed radio data, this enables us to derive a lower limit on the magnetic field
distribution. We stress again that his approach assumes the validity of the hadronic interaction
model. Modelling the CR distribution through different techniques enable
us to bracket our ignorance about the underlying plasma physics that shape the CR distribution and
hence reflect the Bayesian priors that are imposed on the modelling \citep[see][for a
discussion]{article:PinzkePfrommerBergstrom}.

@@

\subsection{Simulation-based approach}
We complement the simplified analytical analysis with a more realistic and predictive approach
derived from cosmological hydrodynamical simulations. We adopt the universal spectral and spatial
gamma-ray model developed by \citet{article:PinzkePfrommer:2010} to estimate the emission from
decaying neutral pions which in clusters dominates over the inverse Compton (IC) emission above
100 MeV. Given a density profile as, e.g., inferred by
cosmological simulations or X-ray observations, the analytic approach models the CR distribution and
the associated radiative emission processes from radio to the gamma-ray band. This formalism was
derived from high-resolution simulations of clusters of galaxies that included radiative hydrodynamics,
star formation, supernova feedback, and followed CR physics by tracing the most important injection
and loss processes self-consistently while accounting for the CR pressure in the equation of motion
\citep{article:PfrommerSpringelEnsslinJubelgas, article:EnsslinPfrommerSpringelJubelgas:2007,
  article:JubelgasSpringelEnsslinPfrommer:2008}. The results are in line with earlier numerical
results on some of the overall characteristics of the CR distribution and the associated radiative
emission processes \citep{article:DolagEnsslin:2000, article:MiniatiRyuKangJones:2001,
  article:Miniati:2003, article:Pfrommer_etal:2007, article:PfrommerEnsslinSpringel:2008,
  article:Pfrommer:2008}.

@@

Figure \ref{fig:spectrum} shows the expected integral spectral energy distribution of Coma within
the virial radius (dotted lines). Additionally shown are integrals of the differential spectrum for
finited energy intervals across the angular apertures tested in this study (solid lines). These
model fluxes are compared to {\em Fermi} and VERITAS flux upper limits of the same energy intervals,
summarized in Table \ref{table:constraints}.  The gamma-ray flux limit of {\em Fermi} in the energy
interval of 1--3 GeV ($<0.4^\circ$) is most constraining but still above the model predictions that
assume an optimistically large shock acceleration efficiency and CR transport parameters as laid
out above.  In our further analysis, we use the most constraining {\em Fermi} flux limits in the
energy interval of 1-3 GeV as well as the gamma-ray flux limits of VERITAS in the energy range above
220 GeV.

Figure \ref{fig:xcr} shows the relative CR pressure, $X_{\CR} =
\expval{P_{\CR}}/\expval{P_\rmn{th}}$ as a function of radial distance, $R$, from the Coma cluster
center (red lines) and contained within $R$ (blue lines). All radii are shown in units of the virial
radius, $R_\rmn{vir}=2.2$~Mpc. To compute the CR pressure, we assume a low-momentum cutoff of the CR
distribution at $q = 0.8\,m_p c$ to reflect tge high cooling rates at low CR energies. The relative CR
pressure rises toward the outer regions on account of the higher efficiency of CR acceleration at the
peripheral accretion shocks compared to the weak central flow shocks. Adiabatic compression of a
mixture of CRs and thermal gas disfavours the
CR pressure relative to the thermal pressure on account of the softer equation of state of CRs. The
weak increase of $X_{\CR}$ toward the core is due to the comparably fast thermal cooling of gas.

@@

\subsection{Constraining the Magnetic Field}
\label{sec:B}
In the previous section we have obtained an absolute lower limit in the gamma-ray emission in
the hadronic model by assuming high magnetic fields, $B\gg B_\rmn{CMB}$. We can turn the
argument around and use our upper limit on the gamma-ray emission (and by extension on the
CR pressure) to infer a lower limit on the magnetic field needed to explain the observed radio
emission. This, again, assumes the validity of the hadronic model of radio halos, in which the
radio-emitting electrons are secondaries from CR interactions. Lowering the gamma-ray
constraint will tighten the magnetic field limit. In case of a conflict with magnetic field
measurements by other methods, e.g., Faraday rotation measure (RM),\footnote{Generally,
Faraday RM analysis of the magnetic field strength by, e.g., background sources through
clusters are degenerate with the magnetic coherence scale and may be biased by the unknown
correlation between magnetic and density fluctuations.} the hadronic model of radio halos would
be challenged. The method we use to constrain the magnetic field inherits a dependence on
the assumed radial scaling which we parameterize as
\begin{equation}
\label{eq:B}
B(r) = B_{0} \,\left(\frac{n_{\rmn{e}}(r)}{n_{\rmn{e}}(0)}\right)^{\alpha_B},
\end{equation}
as suggested by Faraday RM studies and numerical magneto-hydrodynamical (MHD) simulations
\citep[][and references therein]{article:Bonafede_etal:2010, article:Bonafede_etal:2011}. In fact,
the magnetic field in the Coma cluster is among the best constrained, because its proximity permits
RM observations of seven radio-sources located at projected distances of 50 to 1500 kpc from the
cluster center. The best-fit model yields $B_{0} = 4.7^{+0.7}_{-0.8}\,\mu$G and
$\alpha_{B} = 0.5^{+0.2}_{-0.1}$ \citep{article:Bonafede_etal:2010}. We aim for constraining the
central field strength, $B_{0}$, and permit the magnetic decline, $\alpha_{B}$, to vary
within a reasonable range of $\Delta\alpha_{B}=0.2$ as suggested by the Faraday RM studies.
We proceed as follows:

@@

As a final word, in Table~\ref{table:Bmin} we additionally show the corresponding values for the
relative CR pressure (at the largest emission radius at 1~Mpc) such that the model reproduces the
observed radio surface brightness profile. They should be interpreted as upper limits since they
are derived from flux upper limits. For the {\em Fermi} upper limits, they range from 0.1 to 0.4;
hence the $X_{\CR}$-profiles always obey the energy condition, i.e., $P_{\CR} < P_{\mathrm{th}}$
over the entire range of the radio halo emission ($< 1$ Mpc).\footnote{See Figure 3 in
\citet{article:PfrommerEnsslin:2004a} for the entire parameter range assuming minimum energy
conditions and \citet{article:PfrommerEnsslin:2004b}, Figure 7 for a parametrization as adopted in
this study.  We caution, however, that the minimum-energy condition is violated at the outer radio
halo boundary for the range of minimum magnetic-field values inferred by this study.} The
corresponding values for $X_\CR$ in the cluster center are smaller than 0.01 for the entire
parameter space probed in this study. We conclude that the hadronic model is not challenged by
current Faraday RM data and a perfectly viable possibility in explaining the Coma radio halo
emission.

@@

Given the upper limit on the observed gamma-ray rate,
$R_{\gamma}(95\%\ \mathrm{CL}) = N_{\gamma}(95\%\ \mathrm{CL}) / T_{\mathrm{obs}}$, we can place
constraints on the WIMP parameter space $(m_{\chi}, \expval{\sigma v})$. From Eq.
(\ref{eqn:WIMPflux}) we get
\begin{equation}
\expval{\sigma v}(95\%\ \mathrm{CL}) > 
R_{\gamma}(95\%\ \mathrm{CL}) \frac{8\pi m_{\chi}^{2}}{J(\Delta\Omega)}
\frac{1}{\int^{m_{\chi}} dE A_{\mathrm{eff}}dN_{\gamma}/dE}
\end{equation}
Because the self-annihilation of a WIMP is a two-body process, the astrophysical factor
$J(\Delta\Omega)$ can be expressed as the line-of-sight integral of the DM density squared
\begin{equation}
J(\Delta\Omega)=\left(\frac{1}{\rho_{c}^{2}R_{H}}\right)
\int_{\Delta\Omega}d\Omega\int d\lambda \rho^{2}(\lambda,\Omega),
\end{equation}
where $\lambda$ represents the line-of-sight. Here,
$J(\Delta\Omega)$ has been normalized to the square of the critical density, $\rho_{c}$, and the
Hubble radius, $R_{H}$. In this work we have adopted the DM density profile by
Navarro-Frenk-White \citep[NFW;][]{article:NavarroFrenkWhite:1997},
\begin{equation}
\rho_{\chi}(r)=\rho_{s}\left(\frac{r}{r_{s}}\right)^{-1}\left(1+\frac{r}{r_{s}}\right)^{-2},
\end{equation} 
which parametrizes the distribution of DM by the scale radius $r_{s}$ and the scale density
$\rho_{s}$. Using weak-lensing measurements of the virial mass in the Coma cluster
\citep{article:Gavazzi_etal:2009} we find for these parameters $r_{s}=0.654$ Mpc and
$\rho_{s}=4.4\times 10^{4}$ M$_{\odot}$/Mpc$^{3}$. Table \ref{table:astrofactor} lists the
astrophysical factor calculated for the different VERITAS appertures considered in this work.
It also lists the astrophysical factor calculated for the background region, which has been used
to estimate the gamma-ray contamination due to DM in the background region.

The resulting exclusion regions in the $(\expval{\sigma v}, m_{\chi})$ parameter space are shown in
Figure \ref{fig:dm} for three different DM self-annihilation channels: W$^{+}$W$^{-}$,
b$\bar{\mathrm{b}}$, and $\tau^{+}\tau^{-}$. Depending on the chosen apperture the limits are on
the order of $10^{-20}$ to $10^{-21}$ cm$^{3}$ s$^{-1}$. The numerical values for each of the channels
are listed in Table \ref{table:DMlimits}. We note that these are based on conservative estimates of the
astrophysical factor $J$, as we do not assume boosting by clumping or the Sommerfeld effect.

@@

\begin{figure*}
\begin{center}
\scalebox{1.0}{\plotone{f4.eps}}
\end{center}
\caption{{\em Figure TBC} Integral flux upper limits for Coma (this work, Tables~\ref{table:results}
  and \ref{table:fermi}) for different integration radii (shown with different colors), assuming
  $\alpha=2.1$. These are compared to integrated spectra of the same energy interval and aperture,
  assuming the universal gamma-ray spectrum of clusters \citep{article:PinzkePfrommer:2010}. To
  guide the eye, we show the underlying universal integral energy distribution of pion decay
  gamma-rays, $E_\gamma F_\gamma(>E_\gamma)$, resulting from hadronic interactions of CRs and ICM
  protons (dotted). Note that the {\em Fermi} limit for the energy interval of 1--3 GeV within the
  aperture of 0.4$^\circ$ is most constraining.}
\label{fig:spectrum}
\end{figure*}

\begin{figure*}
\begin{center}
\scalebox{1.0}{\plotone{f5.eps}}
\end{center}
\caption{{\em Figure TBC} CR-to-thermal pressure, $X_{\mathrm{CR}}=\expval{P_{\CR}}/ \expval{P_{\mathrm{th}}}$, as a
function of radial distance from the center of the Coma. Model predictions are shown by dashed
lines \citep{article:PinzkePfrommer:2010} while these are scaled to match the most constraining
VERITAS upper limits within 0.2$^\circ$ (solid) and {\em Fermi} upper limits within 0.4$^\circ$
(dashed). We compare differential $X_\CR$-profiles (red) to integrated profiles
$X_{\CR}(<R/R_{\rmn{vir}})=\int_0^R P_\CR\, dV / \int_0^R P_{\rmn{th}}\, dV$ (blue) which we use to
compare to the upper limits.}
\label{fig:xcr}
\end{figure*}

>>>>

\begin{abstract}
  Observations of radio halos and relics in galaxy clusters are indicative of efficient acceleration
  processes of electrons while this process should be even more efficient for protons. In
  combination with their long cooling time which exceeds the Hubble time, this suggests they may also
  be sources of very high-energy (VHE) gamma-ray emission as a result of highly relativistic
  cosmic-ray (CR) protons and ions interacting with intracluster gas. Reported here are VHE
  gamma-ray observations of the Coma galaxy cluster with the VERITAS array of imaging Cherenkov
  telescopes, with complementing Fermi Gamma-ray Space Telescope observations at GeV energies. No
  significant gamma-ray emission from the Coma cluster was detected. Flux upper limits at the 99\%
  confidence level were measured to be on the order of $2-5\times 10^{-8}$ ph. m$^{-2}$ s$^{-1}$
  (VERITAS, $>$220 GeV) and $\sim 3\times 10^{-6}$ ph. m$^{-2}$ s$^{-1}$ ({\em Fermi}, 1-3 GeV),
  respectively. For {\em Fermi} (VERITAS), the limits are a factor of 20 (60) higher than
  model-predicted minimum gamma-ray fluxes, which assume the radio-emitting electrons to be all
  secondaries from cosmic-ray interactions. We use the gamma-ray flux upper limits to constrain the
  populations of CRs and magnetic fields and in Coma. Using an analytical approach informed by
  cosmological cluster simulations, the CR-to-thermal pressure is constrained to be smaller than
  15\% from VERITAS data and smaller than 2.5\% from {\em Fermi} data (averaged within the virial
  radius). Assuming that the radio-emitting electrons of the Coma halo result from hadronic CR
  interactions, the observations imply a lower limit on the intra-cluster magnetic field strengths
  of about $1$--$2\mu$G, depending on the magnetic decline and gamma-ray spectral index.  Since
  these values are well below those inferred by Faraday rotation measurements in Coma, this renders
  the hadronic model a very plausible explanation of the Coma radio halo. Finally, a constraint on
  the thermally averaged product of the total self-annihilation cross section and the relative
  velocity of the dark-matter particles, $\expval{\sigma v}$, is derived and found to be $\lesssim
  10^{-21}$ cm$^{3}$ s$^{-1}$ for $m_{\chi}\gtrsim 1000$ GeV/c$^{2}$.
\end{abstract}

@@

The ICM is a hot, $T\sim 10^{8}$ K, plasma emitting thermal bremsstrahlung in the soft
X-rays regime \citep[see, e.g.,][]{article:Petrosian:2001}. This plasma has been heated
primarily through collisionless structure-formation shocks that form as a result of the
hierarchical merging and accretion processes. Such shocks and turbulence in the ICM gas and
intra-cluster magnetic fields also provide a means to efficiently accelerate particles \citep[see,
e.g.,][]{article:ColafrancescoBlasi:1998, article:Ryu_etal:2003}. Many clusters do indeed feature
mega-parsec scale halos of non-thermal radio emission, indicative of a population of relativistic
electrons and positrons (leptons) and magnetic fields permeating the ICM
\citep{article:Cassano_etal:2010}. The two theories competing to explain the radio halo
are the ``hadronic model'', in which the radio-emitting electrons are produced in inelastic
collisions of cosmic-ray (CR)  ions with thermal gas of the ICM \citep{article:Dennison:1980},
%\citep[][and references therein]{article:EnsslinPfrommerMiniatiSubramanian:2011}
and the ``re-acceleration model'', in which during states of strong ICM turbulence, e.g. after a
cluster merger, interactions of a long-lived pool of 100-MeV electrons with plasma waves might
be efficient enough to accelerate them to energies ($\sim 10$ GeV) sufficient to produce observable
radio emission \citep[][and references therein]{article:BrunettiLazarian:2010}. Recent observations of
possibly non-thermal emission from clusters in the the extreme ultraviolet
\citep[EUV; ][]{article:SarazinLieu:1998} and hard X-rays \citep{article:RephaeliGruber:2002, article:Fusco-Femiano_etal:2004, article:Eckert_etal:2007} may provide further indications of
relativistic particle populations in clusters, although the interpretation of these observations as non-thermal diffuse emission have been disputed on the basis of more sensitive observations
\citep[see, e.g.,][]{article:Ajello_etal:2009, article:Ajello_etal:2010, article:Wik_etal:2009}.

@@

\section{Gamma Ray Emission from Cosmic Rays}
We decided to adopt a multi-faceted approach to constrain the CR-to-thermal pressure distribution
in the Coma cluster using the upper limits derived from VERITAS and \emph{Fermi}-LAT data in this
work. This approach includes (1) a simplified multi-frequency analytical approach that assumes a
constant CR-to-thermal energy density and a power-law spectrum in momentum, (2) an analytic model
derived from cosmological hydrodynamical simulations of the formation of galaxy clusters, and (3)
an approach that uses the
observed luminosity of the giant radio halo in Coma to place a lower limit on the expected
$\gamma$-ray flux in the hadronic model -- where the radio-emitting electrons are secondaries from
CR interactions and which is independent of the magnetic field distribution.  This translates into a
minimum CR pressure which, if challenged by tight gamma-ray limits/detections, enables scrutinizing
the hadronic interaction model of the formation of giant radio halos. Alternatively, by realizing a
spatial CR distribution that is just consistent with the flux upper limits and requiring the model to
match the observed radio data, this enables us to derive a lower limit on the magnetic field
distribution. We stress again that his approach assumes the validity of the hadronic interaction
model. Modelling the CR distribution through different techniques enable
us to bracket our ignorance about the underlying plasma physics that shape the CR distribution and
hence reflect the Bayesian priors that are imposed on the modelling \citep[see][for a
discussion]{article:PinzkePfrommerBergstrom}.

@@

\subsection{Simulation-based approach}
\label{sec:simulation}
We complement the simplified analytical analysis with a more realistic and predictive approach
derived from cosmological hydrodynamical simulations. We adopt the universal spectral and spatial
gamma-ray model developed by \citet{article:PinzkePfrommer:2010} to estimate the emission from
decaying neutral pions which in clusters dominates over the inverse Compton (IC) emission above
100 MeV. Given a density profile as, e.g., inferred by
cosmological simulations or X-ray observations, the analytic approach models the CR distribution and
the associated radiative emission processes from radio to the gamma-ray band. This formalism was
derived from high-resolution simulations of clusters of galaxies that included radiative hydrodynamics,
star formation, supernova feedback, and followed CR physics by tracing the most important injection
and loss processes self-consistently while accounting for the CR pressure in the equation of motion
\citep{article:PfrommerSpringelEnsslinJubelgas, article:EnsslinPfrommerSpringelJubelgas:2007,
  article:JubelgasSpringelEnsslinPfrommer:2008}. The results are in line with earlier numerical
results on some of the overall characteristics of the CR distribution and the associated radiative
emission processes \citep{article:DolagEnsslin:2000, article:MiniatiRyuKangJones:2001,
  article:Miniati:2003, article:Pfrommer_etal:2007, article:PfrommerEnsslinSpringel:2008,
  article:Pfrommer:2008}.

@@

Figure \ref{fig:spectrum} shows the expected integral spectral energy distribution of Coma within
the virial radius (dotted lines). Additionally shown are integrals of the differential spectrum for
finite energy intervals across the angular apertures tested in this study (solid lines). These
model fluxes are compared to {\em Fermi} and VERITAS flux upper limits of the same energy intervals,
summarized in Table \ref{table:constraints}.  The gamma-ray flux limit of {\em Fermi} in the energy
interval of 1--3 GeV ($<0.4^\circ$) is most constraining but still above the model predictions that
assume an optimistically large shock acceleration efficiency and CR transport parameters as laid
out above.  In our further analysis, we use the most constraining {\em Fermi} flux limits in the
energy interval of 1-3 GeV as well as the gamma-ray flux limits of VERITAS in the energy range above
220 GeV.

Figure \ref{fig:xcr} shows the relative CR pressure, $X_{\CR} =
\expval{P_{\CR}}/\expval{P_\rmn{th}}$ as a function of radial distance, $R$, from the Coma cluster
center (grey lines) and contained within $R$ (black lines). All radii are shown in units of the virial
radius, $R_\rmn{vir}=2.2$~Mpc. To compute the CR pressure, we assume a low-momentum cutoff of the CR
distribution at $q = 0.8\,m_p c$ to reflect the high cooling rates at low CR energies. The relative CR
pressure rises toward the outer regions on account of the higher efficiency of CR acceleration at the
peripheral accretion shocks compared to the weak central flow shocks. Adiabatic compression of a
mixture of CRs and thermal gas disfavours the
CR pressure relative to the thermal pressure on account of the softer equation of state of CRs. The
weak increase of $X_{\CR}$ toward the core is due to the comparably fast thermal cooling of gas.

@@

\subsection{Constraining the Magnetic Field}
\label{sec:B}
In the previous section we have obtained an absolute lower limit in the gamma-ray emission in
the hadronic model by assuming high magnetic fields, $B\gg B_\rmn{CMB}$. We can turn the
argument around and use our upper limit on the gamma-ray emission (and by extension on the
CR pressure) to infer a lower limit on the magnetic field needed to explain the observed radio
emission. This, again, assumes the validity of the hadronic model of radio halos, in which the
radio-emitting electrons are secondaries from CR interactions. Lowering the gamma-ray
constraint will tighten the magnetic field limit. In case of a conflict with magnetic field
measurements by other methods, e.g., Faraday rotation measure (RM),\footnote{Generally,
Faraday RM analysis of the magnetic field strength by, e.g., background sources through
clusters are degenerate with the magnetic coherence scale and may be biased by the unknown
correlation between magnetic and density fluctuations.} the hadronic model of radio halos would
be challenged. The method we use to constrain the magnetic field inherits a dependence on
the assumed radial scaling which we parametrize as
\begin{equation}
\label{eq:B}
B(r) = B_{0} \,\left(\frac{n_{\rmn{e}}(r)}{n_{\rmn{e}}(0)}\right)^{\alpha_B},
\end{equation}
as suggested by Faraday RM studies and numerical magneto-hydrodynamical (MHD) simulations
\citep[][and references therein]{article:Bonafede_etal:2010, article:Bonafede_etal:2011}. In fact,
the magnetic field in the Coma cluster is among the best constrained, because its proximity permits
RM observations of seven radio-sources located at projected distances of 50 to 1500 kpc from the
cluster center. The best-fit model yields $B_{0} = 4.7^{+0.7}_{-0.8}\,\mu$G and
$\alpha_{B} = 0.5^{+0.2}_{-0.1}$ \citep{article:Bonafede_etal:2010}. We aim for constraining the
central field strength, $B_{0}$, and permit the magnetic decline, $\alpha_{B}$, to vary
within a reasonable range of $\Delta\alpha_{B}=0.2$ as suggested by the Faraday RM studies.
We proceed as follows:

@@

As a final word, in Table~\ref{table:Bmin} we additionally show the corresponding values for the
relative CR pressure (at the largest emission radius at 1~Mpc) such that the model reproduces the
observed radio surface brightness profile.\footnote{Note that in this section, we determine the
  radial behavior of $X_\CR$ by adopting a specific model for the magnetic field and requiring the
  modeled synchrotron surface brightness profile to match the observed data of the Coma radio
  halo. This is in contrast to the simplified analytical CR model where $X_\CR=\rmn{const.}$
  (Sect.\ref{sec:simple}) and to the simulation-based model where $X_\CR(r)$ is derived from
  cosmological cluster simulations (Sect.\ref{sec:simulation}).} They should be interpreted as upper
limits since they are derived from flux upper limits. For the {\em Fermi} upper limits, they range
from 0.1 to 0.4; hence the $X_{\CR}$-profiles always obey the energy condition, i.e., $P_{\CR} <
P_{\mathrm{th}}$ over the entire range of the radio halo emission ($< 1$ Mpc).\footnote{See Figure 3
  in \citet{article:PfrommerEnsslin:2004a} for the entire parameter range assuming minimum energy
  conditions and \citet{article:PfrommerEnsslin:2004b}, Figure 7 for a parametrization as adopted in
  this study.  We caution, however, that the minimum-energy condition is violated at the outer radio
  halo boundary for the range of minimum magnetic-field values inferred by this study.} The
corresponding values for $X_\CR$ in the cluster center are smaller than 0.01 for the entire
parameter space probed in this study. We conclude that the hadronic model is not challenged by
current Faraday RM data and a perfectly viable possibility in explaining the Coma radio halo
emission.

@@

Given the upper limit on the observed gamma-ray rate,
$R_{\gamma}(95\%\ \mathrm{CL}) = N_{\gamma}(95\%\ \mathrm{CL}) / T_{\mathrm{obs}}$, we can place
constraints on the WIMP parameter space $(m_{\chi}, \expval{\sigma v})$. From Eq.
(\ref{eqn:WIMPflux}) we get
\begin{equation}
\expval{\sigma v}(95\%\ \mathrm{CL}) > 
R_{\gamma}(95\%\ \mathrm{CL}) \frac{8\pi m_{\chi}^{2}}{J(\Delta\Omega)}
\frac{1}{\int^{m_{\chi}} dE A_{\mathrm{eff}}dN_{\gamma}/dE}
\end{equation}
Because the self-annihilation of a WIMP is a two-body process, the astrophysical factor
$J(\Delta\Omega)$ can be expressed as the line-of-sight integral of the DM density squared
\begin{equation}
J(\Delta\Omega)=\left(\frac{1}{\rho_{c}^{2}R_{H}}\right)
\int_{\Delta\Omega}d\Omega\int d\lambda \rho^{2}(\lambda,\Omega),
\end{equation}
where $\lambda$ represents the line-of-sight. Here,
$J(\Delta\Omega)$ has been normalized to the square of the critical density, $\rho_{c}$, and the
Hubble radius, $R_{H}$. In this work we have adopted the DM density profile by
Navarro-Frenk-White \citep[NFW;][]{article:NavarroFrenkWhite:1997},
\begin{equation}
\rho_{\chi}(r)=\rho_{s}\left(\frac{r}{r_{s}}\right)^{-1}\left(1+\frac{r}{r_{s}}\right)^{-2},
\end{equation} 
which parametrizes the distribution of DM by the scale radius $r_{s}$ and the scale density
$\rho_{s}$. Using weak-lensing measurements of the virial mass in the Coma cluster
\citep{article:Gavazzi_etal:2009} we find for these parameters $r_{s}=0.654$ Mpc and
$\rho_{s}=4.4\times 10^{4}$ M$_{\odot}$/Mpc$^{3}$. Table \ref{table:astrofactor} lists the
astrophysical factor calculated for the different VERITAS apertures considered in this work.
It also lists the astrophysical factor calculated for the background region, which has been used
to estimate the gamma-ray contamination due to DM in the background region.

The resulting exclusion regions in the $(\expval{\sigma v}, m_{\chi})$ parameter space are shown in
Figure \ref{fig:dm} for three different DM self-annihilation channels: W$^{+}$W$^{-}$,
b$\bar{\mathrm{b}}$, and $\tau^{+}\tau^{-}$. Depending on the chosen aperture the limits are on
the order of $10^{-20}$ to $10^{-21}$ cm$^{3}$ s$^{-1}$. The numerical values for each of the channels
are listed in Table \ref{table:DMlimits}. We note that these are based on conservative estimates of the
astrophysical factor $J$, as we do not assume boosting by clumping or the Sommerfeld effect.

@@

\begin{figure*}
\begin{center}
\scalebox{1.0}{\plotone{f4.eps}}
\end{center}
\caption{Integral flux upper limits for Coma (this work, Tables~\ref{table:results}
  and \ref{table:fermi}) for different integration radii (shown with different colors), assuming
  $\alpha=2.1$. These are compared to integrated spectra of the same energy interval and aperture,
  assuming the universal gamma-ray spectrum of clusters \citep{article:PinzkePfrommer:2010}. To
  guide the eye, we show the underlying universal integral energy distribution of pion decay
  gamma-rays, $E_\gamma F_\gamma(>E_\gamma)$, resulting from hadronic interactions of CRs and ICM
  protons (dotted). Note that the {\em Fermi} limit for the energy interval of 1--3 GeV within the
  aperture of 0.4$^\circ$ is most constraining.}
\label{fig:spectrum}
\end{figure*}

\begin{figure*}
\begin{center}
\scalebox{1.0}{\plotone{f5.eps}}
\end{center}
\caption{CR-to-thermal pressure, $X_{\mathrm{CR}}=\expval{P_{\CR}}/ \expval{P_{\mathrm{th}}}$, as a
function of radial distance from the center of the Coma. Model predictions are shown by dashed
lines \citep{article:PinzkePfrommer:2010} while these are scaled to match the most constraining
VERITAS upper limits within 0.2$^\circ$ (solid) and {\em Fermi} upper limits within 0.4$^\circ$
(dashed). We compare differential $X_\CR$-profiles (grey) to integrated profiles
$X_{\CR}(<R/R_{\rmn{vir}})=\int_0^R P_\CR\, dV / \int_0^R P_{\rmn{th}}\, dV$ (black) which we use to
compare to the upper limits.}
\label{fig:xcr}
\end{figure*}
